---
title: "ENVS 193DS Final"
author: "Matthew Schoen"
date: June 9, 2026
format: pdf
toc: true
---

Link to ENVS 193DS Final GitHub Repo: [HERE](https://github.com/meschoen76/ENVS-193DS_spring-2026_final)

# Set Up

```{r}
#| label: set-up
#| message: false

library(janitor)
library(tidyverse)
library(here)
library(readxl)
library(flextable)
library(DHARMa)
library(MuMIn)
library(ggeffects)

# Reading in the data & storing it as an object
nest <- read.csv(here("data", "nest_data_final.csv"))
```

# Problem 1: Research Writing 

## a. Transparents Statistical Methods 

**Part 1:**

In part 1, they more than likely used a logistic regression as the response variable is the probably of American Avocet microhabitat use while the predictor variable is the distance to the water's edge. 

**Part 2:**

In part 2, the they more than likely used a chi-square test of independence as both bird species and habitat type are categorical variables and the analysis looks to examine how the distribution of bird species differs among habitat type. 

## b. Suggestions for rewriting 

**Part 1:** 

American Avocet tuse of micro-habtitats varies wiht distance from the water edge. This suggests that proximity to water influences where and how the American Avocet chooses to use habitats.


**Part 2:**

American Avocets, Forster's Terns, and Black-necked Stilts used habitat types differently. This suggests that each species has habitat preference differs within the San Francisco Bay estuary. 

## c. Further Analysis

To further the analysis, examining how the type of vegetation cover within a given area influences American Avoocet habitat use. This *continuous variable* could be measured as a percentage (%) of ground cover within the study area using quadrats set at a given dimension (ex: 1m x 1m, 3ft x 3ft) and could showcase how other factors affected by vegetation cover (food, nesting area, protection from predators) influences behavior.

# Problem 2: Data Analysis

**Goal:** Answer the following questions 

1) How do tree height and acacia ant species influence the probability of bird nest occurrence (of any species of birds)?

2) Does acacia ant species identity influence the effect of tree height on probability of nest occurrence differ depending on acacia ant species?

## a. Clean & Wrangle 

```{r}
#| label: cleaning-and-wrangling

# Creating new clean object from nest
nests_clean <- nest |>  
  # selecting only the necessary columns 
  select("case_control", # selecting case_control
         "height_cm", # selecting bird_speces
         "ant_species") |>  # selecting ant_species
  # filtering ant_species for the specified species
  filter(ant_species %in% c("RRB", "AB", "BBR")) |> 
  # reordering ant species as specied 
  mutate(ant_species = 
         fct_relevel(ant_species,
           "AB",
           "RRB",
           "BBR")) |> 
  # Creating new column with spelled out names 
 mutate(ant_species,
        ant_species_name = case_when( # specifying specific names in the new column 
    ant_species == "AB" ~ "Crematogaster sjostedti",
    ant_species == "RRB" ~ "Crematogaster mimosae",
    ant_species == "BBR" ~ "Crematogaster nigriceps"
  ))

# Displaying structure of nests_clean
str(nests_clean)

# Displaying 10 random rows within the data set 
slice_sample(nests_clean, n = 10)
```

## b. Response Variable 

In this data set, "case_control" is made up of a binary system where 1 indicates a bird nest is present in a tree while 0 indicates no bird nests were found within a tree. This is important when assessing how/why birds choose specific trees for nesting. 

## c. Purpose of the Study 

**Tree Height:**

Tree height is a continuous variable measured in centimeters. As tree height increases, the probability of nest occurrence increases as a taller tree provides greater protection from predators, visibility, and increases structural support to the nest itself (Found in the "Introduction" section).

**Ant Species:**

Ant species is a discrete categorical variable as it contains three distinct "categories" which in this case are the different species of ants: *Crematogaster sjostedti*, *Crematoogaster mimosae*, and *Crematoogaster nigriceps*. The species of ant can influence nesting probability as different species of ants vary in their overall "aggressiveness" toward other organsisms which in turn alters tree structure and can affect the overall suitability of a tree for nesting. (Found in Introduction & Methods).

## d. Table of Models 





## e. Run the Models 

## f. Select the Best Model 

## g. Check the Diagnostics 

## h. Visualize the Model Predictions 

## i. Write A Figure Caption 

## j. Calculate Model Predictions 

## k. Interpret Your Results 

# Problem 3: Affective and Exploratory Visualizations 

## a. Comparing Visualizations 
## b. Designing An Analysis 

## c. Check Any Assumptions and Run Your Analysis 

## d. Create a Visualization 

## e. Write a Caption 

## f. Write About Your Results 

## g. Sharing Your Affective Visualization 

Completed during week 10 in workshop! 








